|

### 2021-01-14 article_format/article_format_revision.Rmd
Started major changes for resubmission

---
title: |
  A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
#type: ARTICLE TEMPLATE
author:
  - name: Henrik Kenneth Andersen
    affil: a
    email: henrik.andersen@soziologie.tu-chemnitz.de
affiliation:
  - num: a
    address: |
      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
bibliography: references2.bib
# appendix: appendix.tex
abstract: |
  This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extensions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model.
keywords: |
  Fixed effects, structural equation modeling, lavaan, R, panel analysis
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{booktabs}
  \usepackage{bm}
  \usepackage{mathtools}
  \usepackage{amssymb}
  \usepackage{amsmath}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \usepackage[nofiglist]{endfloat}
  \usepackage{blkarray}
  \usepackage{setspace}
  \usepackage{etoolbox}
  \DeclareMathOperator{\E}{\mathbb{E}}
  \DeclareMathOperator{\Var}{\mathrm{Var}}
  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
  \DeclareMathOperator*{\argmax}{arg\,max}
  \DeclareMathOperator*{\argmin}{arg\,min}
  \mathtoolsset{showonlyrefs}
  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
  \def\tightlist{}
  \pagenumbering{gobble}
output: rticles::tf_article
---

@@

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable. Fixed effects (FE) regression involves controlling for these time-invariant influences via a number of various methods. It thus accounts for a likely and common source of bias.

Structural equation modeling (SEM) is a popular regression framework. One of its main strengths is its flexibility. Not only can complex causal structures with multiple dependent variables be tested simultaneously, but in longitudinal (and, more generally, hierarchical) studies both time-varying and invariant predictors can be included, and effects can easily be allowed to vary over time. Thus researchers can allow for and study effects that increase or fade over time, or that appear only in specific periods. Beyond that, with the use of latent variables, SEM provides a way to deal with measurement error and get closer to the true underlying constructs of interest.

There are a number of articles describing basic concept of panel model regression, and FE regression in SEM [e.g., @Allison2011; @Bollen2010; @Teachman2001]. This article is intended as a \textit{practical guide} for researchers looking for in-depth help with specifying FE models in SEM. It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base]. While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example), the `lavaan` package has many benefits. First, like `R` it is open source and completely free. For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them. Second, the implementation of `lavaan` in the larger `R` environment is an enormous advantage. Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object. This means that all aspects of the model, from fit indices, to coefficients and standard errors, to the model matrices, can be accessed and easily integrated into tables and plots. Furthermore, `R` can be used for a great deal of applications. It can be used to manage and manipulate as well as simulate data, perform symbolic algebra, run more traditional analyses (e.g., multiple regression, logistic regression, principal component analysis), etc. Once one is comfortable using `R`, there is no longer any need to switch between different software for data preparation and analysis.

The following article outlines the basic idea of panel regression, the particularities of panel regression SEM, and shows its implementation in `lavaan`. The focus on FE panel regression stems from the fact that the assumptions needed to justify other common models, e.g., random effects, are often implausible. As we will see, the practical differences between the two are, in any event, very small. Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. One of the main strengths of panel SEM compared to the more traditional methods of panel analysis is its flexibility. Therefore, a number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).

# Panel models {#panel}

To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align}
where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$. $\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.

@@

Somewhat unintuitively, the modern approach to FE regression treats the unobserved effect $\alpha_{i}$ as a \textit{random} variable, rather than a fixed parameter to be estimated for each of the cross sectional units [@Wooldridge2002]. And in fact, this view carries over to the way in which FE regression is applied in SEM. So, in order to help facilitate the discussion on the application of FE-SEM later on, it makes sense to spend some time discussing the basic idea of random effects regression (again, as we will see, the practical difference between the two models is very small).

@@

Estimation of the RE model can be done using feasible generalized least squares (GLS) in which the two unknowns in $\bm{\Omega}$, $\sigma_{\alpha}^{2}$ and $\sigma_{\nu}^{2}$, are first estimated using pooled ordinary least squares (pooled OLS or POLS), where^[Normally a degrees-of-freedom correction is applied by subtracting off the number of independent variables that is negligible in large N samples. It is ignored here for the sake of simplicity.]
\begin{align}
\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
\end{align}
and $\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}$, $\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}$ and $\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}$. Then, the coefficients are estimated using those estimates in the variance matrix $\hat{\bm{\Omega}}$:
\begin{align}
\begin{pmatrix}
\hat{\bm{\beta}}_{RE} \\
\hat{\bm{\gamma}}_{RE}
\end{pmatrix} & = 
(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
\end{align}
where $\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}$ and $\bm{X}$ is $NT \times K$ and $\bm{Z}$ is $NT \times M$ and $\bm{y}$ is $NT \times 1$.

@@

Now, we run the FE-SEM in `lavaan`.

```{r}
fe_sem <- '
# Define individual effects variable 
a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
# Regressions, constrain coefficient to be equal over time
y1 ~ b*x1
y2 ~ b*x2 
y3 ~ b*x3
y4 ~ b*x4
y5 ~ b*x5
# Allow unrestricted correlation between eta and covariates
a ~~ x1 + x2 + x3 + x4 + x5
x1 ~~ x2 + x3 + x4 + x5
x2 ~~ x3 + x4 + x5
x3 ~~ x4 + x5
x4 ~~ x5
# Constrain residual variances to be equal over time
y1 ~~ e*y1
y2 ~~ e*y2
y3 ~~ e*y3
y4 ~~ e*y4
y5 ~~ e*y5
'
fe_sem.fit <- sem(model = fe_sem, 
                  data = dfw, 
                  estimator = "ML")
```

@@

>>>>

---
title: |
  A closer look at random and fixed effects panel regression in structural equation modeling using \texttt{lavaan}
type: TEACHER'S CORNER
author:
  - name: Henrik Kenneth Andersen
    affil: a
    email: henrik.andersen@soziologie.tu-chemnitz.de
affiliation:
  - num: a
    address: |
      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
bibliography: references2.bib
# appendix: appendix.tex
abstract: |
  This article provides an in-depth look at random and fixed effects panel regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. 
keywords: |
  Fixed effects, structural equation modeling, lavaan, R, panel regression, longitudinal data 
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{booktabs}
  \usepackage{bm}
  \usepackage{mathtools}
  \usepackage{amssymb}
  \usepackage{amsmath}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \usepackage[nofiglist]{endfloat}
  \usepackage{blkarray}
  \usepackage{setspace}
  \usepackage{etoolbox}
  \DeclareMathOperator{\E}{\mathbb{E}}
  \DeclareMathOperator{\Var}{\mathrm{Var}}
  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
  \DeclareMathOperator*{\argmax}{arg\,max}
  \DeclareMathOperator*{\argmin}{arg\,min}
  \mathtoolsset{showonlyrefs}
  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
  \def\tightlist{}
  \pagenumbering{gobble}
output: rticles::tf_article
---

@@

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable. Random effects (RE) and fixed effects (FE) regression involve accounting for these often unobserved time-invariant influences via a number of methods. In the case of RE models, it is assumed that the stable characteristics are unrelated to the other model covariates. In that case, the stable characteristics, or individual effects as they are often referred to as, will not affect point estimates but are a source of error variance that, if ignored, lead to larger standard errors and increase the likelihood of type II errors (failure to reject a false null hypothesis). FE models are useful when the individual effects are expected to be related to one or more of the model covariates. In that case, these unobserved stable characteristics will act as confounders and lead to biased point estimates. In controlling for these individual effects, FE regression thus accounts for a likely and common source of bias.

Structural equation modeling (SEM) is a popular regression framework. One of its main strengths is its flexibility. Not only can complex causal structures with multiple dependent variables be tested simultaneously, but in longitudinal (and, more generally, hierarchical) studies both time-varying and invariant predictors can be included, and effects can easily be allowed to vary over time. Thus researchers can allow for and study effects that increase or fade over time, or that appear only in specific periods. Beyond that, with the use of latent variables, SEM provides a way to deal with measurement error and get closer to the theoretical constructs of interest.

There are a number of articles describing basic concept of panel model regression, including RE and FE regression in SEM [e.g., @Allison2011; @Bollen2010; @Teachman2001]. This article is intended as a \textit{practical guide} for researchers looking for help specifying panel regression models in SEM. It assumes some basic knowledge of SEM (i.e., minimizing the difference between observed and model-implied covariance matrices and mean vectors, maximum likelihood estimation, etc.);^[For an introduction to SEM, @Bollen1989 is a classic for good reason, and @Ferron2007 lay out exactly how SEM works in one easy to follow article.] perhaps the reader is an experienced SEM-user looking for a quick guide to estimating the static panel models discussed here; or, perhaps the reader is new to SEM, potentially because an adviser or superior has suggested it to them. For the latter case, this article reviews the assumptions associated with the more tradititonal least squares-based approach to RE and FE models, so that the implementation in SEM is more easily to follow. The article focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base]. While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example), the `lavaan` package has many benefits. First, like `R` it is open source and completely free. For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them. Second, the implementation of `lavaan` in the larger `R` environment is an enormous advantage. Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object. This means that all aspects of the model, from fit indices, to coefficients and standard errors, to the model matrices, can be accessed and easily integrated into tables and plots. Furthermore, `R` can be used for a great deal of applications. It can be used to manage and manipulate as well as simulate data, perform symbolic algebra, run more traditional analyses (e.g., multiple regression, logistic regression, principal component analysis), etc. Once one is comfortable using `R`, there is no longer any need to switch between different software for data preparation and analysis.

The following article outlines the basic idea of panel regression, the particularities of panel regression SEM, and shows its implementation in `lavaan`. Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. One of the main strengths of panel SEM compared to the more traditional methods of panel analysis is its flexibility. Therefore, a number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).

# Random and Fixed Effects Panel models {#panel}

It is typically the case that the values for a given unit on a variable at one point in time will tend to tell us something about that unit's values at another point in time. There are two main explanations for this 'empirical regularity' [@Heckman1981; @Hsiao2014, p. 261; @Bruederl2015; @Bianconcini2018]. First, it could be that an experience at one point in time has a causal^[Here the word causal is used to mean 'non-spurious'.] effect on future experiences. In other words, experiencing an event could change the probability of the same or a similar event taking place in the future. For example, if employment increases wages, then the incentive to continue working should increase over time, thereby making it more likely that someone who was employed at one point in time will continue to be employed in the future [@Heckman1981]. When past experiences causally impact future events, the empirical regularity is referred to as 'state dependence'. So-called 'dynamic' panel models that include the lagged dependent variable in the equation for the current dependent variable, like autoregressive and autoregressive cross-lagged models, are examples of panel models that account for state dependence. The second explanation is that correlations over time are due to stable unobservables, like sex, place of birth, motivation, ability or personality characteristics.^[From here on out the units of interest are assumed to be individuals, but other units of observation like schools, companies, countries, etc., are possible.] In other words, stable unit-specific characteristics might predispose individuals to experiencing events with a certain likelihood over time. For example, stable characteristics like an individual's sex and motivation could be part of the reason why some individuals tend to be continuously employed, while others experience spells of unemployment, or are habitually unemployed. This second source of empirical regularity is referred to as individual heterogeneity or, more generally, unobserved heterogeneity [@Wooldridge2012]. So-called 'static' panel models (that do not include the lagged dependent variable in the equation for the current one) focus on this second source of empirical regularity.

The random and fixed effects models that will be discussed here are examples of models that attempt to account for unobserved heterogeneity as a source of empirical regularity. In the case of both the random and fixed effects models, it is assumed that there exist stable individual characteristics that affect the dependent variable of interest over time. Failure to account for this source of stable variance will reduce model fit. If any of the stable characteristics are related to the model covariates (a plausible scenario in many cases where the independent variable(s) of interest are not fixed), then they must also be considered confounders, and failure to account for them will further lead to biased point estimates.^[Whether or not the stable individual characteristics are related to the model covariates is the modern differentiator between random effects, where the individual effects are assumed independent of the covariates, and fixed effects, where the individual effects are assumed to be related to one or more of them, see @Wooldridge2012, p. 251. This can be a cause for confusion for those who are more familiar with the mixed or multilevel model framework. There, 'fixed' effects are those coefficients that apply to the entire sample, 'random' effects are the variance components for the higher level grouping variables, see for example @Hox2010; @Bates2015; @R-lme4.]

In this section, the basic random and fixed effects models, as well as their implementation in SEM, will be discussed. Later on in the article, it will be discussed how to extend the model, by relaxing some assumptions and introducing other time-invariant predictors in a type of 'hybrid' random and fixed effects model. The outlook will describe further extensions, such as the inclusion of lagged dependent variable for a dynamic panel model with individual effects.    
Let us begin with a simple linear 'unobserved effects model' [@Wooldridge2012; @R-plm_a]
\begin{align}
y_{it} & = \beta_{0} + \beta_{1} x_{it} + \alpha_{i} + \varepsilon_{it}, \ i = 1, \ldots, N, \ t = 1, \ldots, T \label{eq:panel-model}
\end{align}
where $y_{it}$ is the dependent variable for unit $i$ at time $t$ and $x_{it}$ is a single covariate with associated scalar coefficient $\beta_{1}$. $\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error. This model could of course be extended by turning $x_{it}$ into a vector of covariates $\bm{x}_{it} = (x_{1_{it}}, x_{2_{it}}, \ldots, x_{k_{it}})$, or by introducing time-invariant covariates, $\bm{z}_{i} = (z_{1_{i}}, z_{2_{i}}, \ldots, z_{m_{i}})$, but we want to first focus on assumptions and mechanics before moving on to discussing possibilities for extending the model.

Equation \eqref{eq:panel-model} is often written in matrix form for unit $i$ as 
\begin{align}
\bm{y}_{i} & = \bm{\iota}_{T}\beta_{0} + \bm{x}_{i}\beta_{1} + \bm{\iota}_{T} \alpha_{i} + \bm{\varepsilon}_{i}, \ i = 1, \ldots, N \label{eq:matrix-panel-model}
\end{align}
where $\bm{y}_{i} = (y_{i1}, y_{i2}, \ldots, y_{iT})^{\intercal}$ and $\bm{x}_{i} = (x_{i1}, x_{i2}, \ldots, x_{iT})^{\intercal}$ are $T \times 1$ vectors of each of the $t$ observations for unit $i$, $\bm{\iota}_{T}$ is a $T \times 1$ vector of ones and $\bm{\varepsilon}_{i}  = (\varepsilon_{i1}, \varepsilon_{i2}, \ldots, \varepsilon_{iT})^{\intercal}$ is likewise a $T \times 1$ vector of errors. It may be helpful to keep this Equation \eqref{eq:matrix-panel-model} in mind later when we restate the model for estimation using SEM. There, we will take the long-format data and transform it into wide-format.

## Assumptions in least squares-based approaches

One of the main strengths of SEM is the flexibility it affords. Nearly every part of the model is specified 'by hand', so to speak (although modern SEM software has come a long way from, say, `EQS` [@Bentler2006], where the errors and disturbances had to be mentioned in the syntax and now comes with a variety of default settings for ease of use), which means model assumptions are conveyed directly, by the user specifying paths and correlations and setting other relations to zero. Common least squares-based statistical packages [like the `plm` package in `R`, @R-plm_a] require less input from the user (usually it is enough to input the model variables and specify some top-level options). The assumptions in the least squares-based models are widely known [@Wooldridge2012; @Bruederl2015; @Greene2012; @Angrist2009] but do not come into play prominently while specifying the model. So here we will discuss the 'typical' assumptions associated with least squares-based random and fixed effects models. Later, when we examine the implementation in SEM, we will then be able to relate back to this discussion.^[What constitutes a 'typical' least squares-based approach to these panel regression models is likely debatable. Here, this refers broadly to the assumptions discussed in the literature and default settings one encounters in statistical packages intended for this purpose.]

We want to relate the SEM-based approach to random and fixed effects to the least squares-based approach which is likely more widely known and more straightforward to implement, i.e., statistical packages for static panel regression models exist (e.g., the `plm` package in `R`, see @R-plm_a), and typically only require the user to enter the model variables and specify some top-level options (random effects vs. fixed effects, individual vs. individual and time effects, etc.). The model assumptions for random and fixed effects models are widely known [@Wooldridge2012; @Bruederl2015; @Greene2012; @Angrist2009] but are not prominent in th

The least squares-based approach to such a panel model typically makes the following base assumptions:^[What constitutes a 'typical' least squares-based approach to these panel regression models is likely debatable. Here, this refers broadly to the default assumptions and settings one encounters in statistical packages intended for this purpose, like the `plm` package in `R`, see @R-plm_a.] first, the mean function is linear, i.e., $\E(y_{it} | x_{it}, \alpha_{i}) = \beta_{0} + \beta_{1} x_{it} + \alpha_{i}$, and second, that the cross-sectional observations are independent, which is assured by random sampling [@Wooldridge2012]. Further, and just as with any other regression model, the idiosyncratic error is assumed to be mean independent of the model covariates and the individual effects, i.e., $\E(\varepsilon_{it} | x_{it}, \alpha_{i}) = \E(\varepsilon_{it}) = 0$.^[If there is an overall intercept, $\beta_{0}$, then it is safe to assume the unconditional expectation of the idiosyncratic error is zero, i.e., $\E(\varepsilon_{it}) = 0$.] The mean independence assumption also implies the assumption of a zero correlation between the idiosyncratic error and the model covariates, i.e., $\E(x_{it}\varepsilon_{it}) = 0$.

The (modern) difference between random and fixed effects boils down to the assumption of the relatedness of the model covariates and the individual effects. The random effects model assumes the model covariates are unrelated to the individual effects, i.e., in this case $\E(\alpha_{i} | x_{it}) = \E(\alpha_{i})$ and thus $\E(\alpha_{i}x_{it}) = 0$. As such, we could put $\alpha_{i}$ into the error term to form the composite error, $\nu_{it} = \alpha_{i} + \varepsilon_{it}$, and the point estimates of the coefficients of interest, here $\beta_{1}$, would be consistent (assuming the model is otherwise correctly specified). The problem with putting the individual effects into the error term is that it would induce dependency between the compostie errors within the same individual, which goes against the independence assumption [@Wooldridge2012; @Schmidheiny2019]. We will discuss that in more detail below. In the fixed effects model, one or more of the model covariates are assumed to be related to the individual effects, here $\E(\alpha_{i} | x_{it}) \ne \E(\alpha_{i})$ and thus $\E(\alpha_{i}x_{it}) \ne 0$. Thus, ignoring the individual heterogeneity and putting $\alpha_{i}$ into the error term would mean that point estimates of the coefficients of interest would be inconsistent.

The traditional least squares-based approaches to random and fixed effects often assume as well that the errors are homoskedastictic over time, i.e., $\E(\varepsilon_{it}^{2}) = \sigma_{\varepsilon_{t}}^{2} = \sigma_{\varepsilon}^{2}$. The individual effects are assumed to have a constant partial effect on the dependent variable over time [@Wooldridge2012, p. 248]. Also, typically least squares-based approaches stack each of the $N$ repeated measures per individual $i$ and calculate one single coefficient per variable for the entire sample. I.e., it is generally not intended that the estimated coefficient should vary across $t$ or $i$.

# Random and fixed effects in SEM

For panel regression in SEM, we typically convert any stacked, long-format vectors of observations of length $NT$ into $T$ separate vectors of length $N$. In doing so, we get $T$ individual equations for each time point
\begin{align}
\bm{y}_{t} & = \bm{\iota}_{N}\beta_{0_{t}} + \bm{x}_{t}\beta_{1_{t}} + \bm{\alpha} + \bm{\varepsilon}_{t}, \ t = 1, \ldots, T \\
\begin{bmatrix}
y_{1_{t}} \\
y_{2_{t}} \\
\vdots \\
y_{N_{t}}
\end{bmatrix} & = 
\begin{bmatrix}
1 \\
1 \\ 
\vdots \\
1
\end{bmatrix} \beta_{0_{t}} + 
\begin{bmatrix}
x_{1_{t}} \\
x_{2_{t}} \\
\vdots \\
x_{N_{t}}
\end{bmatrix} \beta_{1_{t}} + 
\begin{bmatrix}
\alpha_{1} \\
\alpha_{2} \\
\vdots \\
\alpha_{N}
\end{bmatrix} + 
\begin{bmatrix}
\varepsilon_{1_{t}} \\
\varepsilon_{2_{t}} \\
\vdots \\
\varepsilon_{N_{t}}
\end{bmatrix}
\end{align}

Let us discuss the random and fixed effects models in SEM by examining a path diagram. Figure \ref{fig:panel-sem} shows a four-wave panel model with individual effects. As usual, rectangles indicate observed variables and circles indicate latent variables. One-sided errors indicate effects and two-headed errors indicate correlations or covariances.

We assume all of the observed variables are grand mean centered, so that we can drop the overall intercept from the model.

\begin{figure}
\begin{center}
\resizebox{0.75\textwidth}{!}{%
\begin{tikzpicture}
% Node styles ---
\tikzstyle{man} = [rectangle, thick, minimum size = 0.5cm, draw = black!100, fill = white!100, font = \sffamily] 
\tikzstyle{lat} = [circle, thick, minimum size = 0.5cm, draw = black!100, fill = white!100, font = \sffamily] 
\tikzstyle{err} = [rectangle, minimum size = 0.5cm, draw = white!100, fill = white!100, font = \sffamily] 
\tikzstyle{pha} = [rectangle, minimum size = 0.001cm, draw = white, fill = white]
\tikzstyle{con} = [-latex, draw = black!100, font = \sffamily] 
\tikzstyle{seqex} = [latex-latex, draw = black!100, font = \sffamily, dashed]
\tikzstyle{cons} = [-latex, draw = black!100, font = \sffamily\small] 
\tikzstyle{cor} = [latex-latex, font = \sffamily] 
% Begin Figure ---
% Nodes 
\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
\node at (+8.0,+0.0) [man] (x4) {$x_{4}$};
\node at (+4.0,+0.0) [lat] (alp) {$\alpha$};
\node at (+0.0,+2.1) [man] (y1) {$y_{1}$};
\node at (+2.0,+2.1) [man] (y2) {$y_{2}$};
\node at (+6.0,+2.1) [man] (y3) {$y_{3}$};
\node at (+8.0,+2.1) [man] (y4) {$y_{4}$};
\node at (+0.0,+3.0) [err] (e1) {\footnotesize $\varepsilon_{1}$};
\node at (+2.0,+3.0) [err] (e2) {\footnotesize $\varepsilon_{2}$};
\node at (+6.0,+3.0) [err] (e3) {\footnotesize $\varepsilon_{3}$};
\node at (+8.0,+3.0) [err] (e4) {\footnotesize $\varepsilon_{4}$};
% Paths
\path (alp) edge [con] node [near start, fill=white] {1} (y1);
\path (alp) edge [con] node [midway, fill=white] {1} (y2);
\path (alp) edge [con] node [midway, fill=white] {1} (y3);
\path (alp) edge [con] node [near start, fill=white] {1} (y4);
\path (x1) edge [con] node [near start, fill=white] {$\beta$} (y1);
\path (x2) edge [con] node [near start, fill=white] {$\beta$} (y2);
\path (x3) edge [con] node [near start, fill=white] {$\beta$} (y3);
\path (x4) edge [con] node [near start, fill=white] {$\beta$} (y4);
\path (e1) edge [con] node {} (y1);
\path (e2) edge [con] node {} (y2);
\path (e3) edge [con] node {} (y3);
\path (e4) edge [con] node {} (y4);
% correlations
\path (x1) edge [cor, bend right] node {} (x2);
\path (x1) edge [seqex, bend right] node {} (alp);
\path (x1) edge [cor, bend right] node {} (x3);
\path (x1) edge [cor, bend right] node {} (x4);
\path (x2) edge [seqex, bend right] node {} (alp);
\path (x2) edge [cor, bend right] node {} (x3);
\path (x2) edge [cor, bend right] node {} (x4);
\path (x3) edge [seqex, bend left] node {} (alp);
\path (x4) edge [seqex, bend left] node {} (alp); 
\end{tikzpicture}
}
\caption{Four-wave static panel model\label{fig:panel-sem}}
\end{center}
\end{figure}

To begin, let us review a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align}
where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$. $\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.

@@

@@

Estimation of the RE model can be done using feasible generalized least squares (GLS) in which the two unknowns in $\bm{\Omega}$, $\sigma_{\alpha}^{2}$ and $\sigma_{\nu}^{2}$, are first estimated using pooled ordinary least squares (pooled OLS or POLS), where^[Normally a degrees-of-freedom correction is applied by subtracting off the number of independent variables that is negligible in large N samples. It is ignored here for the sake of simplicity.]
\begin{align}
\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
\end{align}
and $\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}$, $\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}$ and $\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}$. Here, $\bm{\hat{\beta}}_{\tiny{POLS}}$ and $\hat{\bm{\gamma}}_{\tiny{POLS}}$ are the POLS estimates for the coefficients of the time-varying and -invariant covariates, respectively. Then, the coefficients are estimated using those estimates in the variance matrix $\hat{\bm{\Omega}}$:
\begin{align}
\begin{pmatrix}
\hat{\bm{\beta}}_{RE} \\
\hat{\bm{\gamma}}_{RE}
\end{pmatrix} & = 
(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
\end{align}
where $\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}$ and $\bm{X}$ is $NT \times K$ and $\bm{Z}$ is $NT \times M$ and $\bm{y}$ is $NT \times 1$.

@@

@@

# Extensions {#extensions}

## Relaxing assumptions meant to mimic traditional FE models {#relax}

There are a number of implicit assumptions attached to the typical FE model that can be relaxed in SEM. Some of these assumptions have been discussed already, and a fairly comprehenisve list of assumptions can be found in @Bollen2010. Here, I will go over just a few, concentrating on the implementation in `lavaan` and the opportunity to empirically test whether the adjustments are justified or not.

The assumptions we will discuss here pertain to the time-invariance of the effects of both the latent individual effects and the observed covariates, as well as a time-invariant error variance. We can also empirically test the correlation between the individual effects and the covariates to see whether a RE model is preferable to the FE model.
